---
title: "steps data"
author: "Cristian"
date: "02. 04. 2016"
output: html_document
---

### Import the dataset
```{r, echo = TRUE}
Sys.setlocale("LC_TIME", "en_GB.UTF-8")
library(knitr)
setwd("/home/cristian/Documents/PhD/lectures/coursera/reproducibleResearch/week1/")
steps <- read.csv("activity.csv")
head(steps)
```

### Transform the dataset
```{r, echo = TRUE}
library(dplyr)
library(lubridate)
steps$day <- as.numeric(steps$date)
stepsDt <- tbl_df(steps)
stepsDay <- group_by(stepsDt, day)
stepsDayTotal <- summarize(stepsDay, sumSteps = sum(steps))
hist(stepsDayTotal$sumSteps, xlab = "Total number of steps / day", main = 
    "Distribution of the number of steps taken per day")
meanSteps <- mean(stepsDayTotal$sumSteps, na.rm = TRUE)
medianSteps <- median(stepsDayTotal$sumSteps, na.rm = TRUE)
```
The mean number of steps taken each day is `r meanSteps`.
The median number of steps taken each day is `r medianSteps`.

### What is the average daily activity pattern?
```{r, echo = TRUE}
stepsInterval <- group_by(stepsDt, interval)
stepsIntervalMean <- summarize(stepsInterval, intervalMean = mean(steps, na.rm = TRUE))
stepsIntervalMean
plot(stepsIntervalMean$interval, stepsIntervalMean$intervalMean, type = "l", 
     xlab = "5-minutes interval", ylab = "Mean number of steps in the interval
     across all days", main = "Time-series plot of the mean number of steps taken
     in each 5-minutes interval on average across all days")
stepsMax <- max(stepsIntervalMean$intervalMean)
intervalMax <- stepsIntervalMean$interval[stepsIntervalMean$intervalMean == stepsMax]
```
The 5-minutes interval `r intervalMax` has the highest average number of steps with `r stepsMax` steps taken in that interval on average.

### Imputing missing values
```{r, echo = TRUE}
stepsNas <- length(steps$steps[is.na(steps$steps)])
nasIndices <- which(is.na(steps$steps))
for (i in nasIndices) {
    steps$steps[i] <- stepsIntervalMean$intervalMean[stepsIntervalMean$interval == steps$interval[i]]
}
stepsDt <- tbl_df(steps)
stepsDt
```
There are `r stepsNas` intervals for which the number of steps taken is not available. Each missing value is replaced by the mean number of steps taken during that interval across all days.
Check that the imputation has worked properly by plotting the time-series plot again. It should be the same as before because the mean in each interval should have remained the same.
```{r, echo = TRUE}
stepsInterval <- group_by(stepsDt, interval)
stepsIntervalMean <- summarize(stepsInterval, intervalMean = mean(steps, na.rm = TRUE))
stepsIntervalMean
plot(stepsIntervalMean$interval, stepsIntervalMean$intervalMean, type = "l", 
     xlab = "5-minutes interval", ylab = "Mean number of steps in the interval
     across all days", main = "Time-series plot of the mean number of steps taken
     in each 5-minutes interval on average across all days")
```
The number of NAs is now `r length(steps$steps[is.na(steps$steps)])`.

```{r, echo = TRUE}
stepsDayComplete <- group_by(stepsDt, day)
stepsDayCompleteTotal <- summarize(stepsDayComplete, sumStepsComplete = sum(steps))
hist(stepsDayCompleteTotal$sumStepsComplete, xlab = "Total number of steps / day", main = 
    "Distribution of the number of steps taken per day")
meanStepsComplete <- mean(stepsDayCompleteTotal$sumStepsComplete, na.rm = TRUE)
medianStepsComplete <- median(stepsDayCompleteTotal$sumSteps, na.rm = TRUE)
```
The mean number of steps taken each day is now `r meanStepsComplete`, unchanged from before.
The median number of steps taken each day is `r medianStepsComplete`, slightly higher than before.

### Are there differences in activity patterns between weekdays and weekends?
```{r, echo = TRUE}
steps$weekday <- weekdays(as.Date(steps$date))
steps$dayType <- rep("weekday", nrow(steps))
steps$dayType[steps$weekday == "Saturday" | steps$weekday == "Sunday"] <- "weekend"
par(mfrow = c(2, 1))
# Weekdays
stepsWeekdays <- tbl_df(steps[steps$dayType == "weekday", ])
stepsWeekdaysInterval <- group_by(stepsWeekdays, interval)
stepsWeekdaysIntervalMean <- summarize(stepsWeekdaysInterval, intervalMean = mean(steps, na.rm = TRUE))
stepsWeekdaysIntervalMean
plot(stepsWeekdaysIntervalMean$interval, stepsWeekdaysIntervalMean$intervalMean, 
     type = "l", xlab = "5-minutes interval", ylab = "Mean number of steps in the interval
     across all weekdays", main = "Time-series plot of the mean number of steps taken
     in each 5-minutes interval on average across all weekdays")
# Weekends
stepsWeekends <- tbl_df(steps[steps$dayType == "weekend", ])
stepsWeekendsInterval <- group_by(stepsWeekends, interval)
stepsWeekendsIntervalMean <- summarize(stepsWeekendsInterval, intervalMean = mean(steps, na.rm = TRUE))
stepsWeekendsIntervalMean
plot(stepsWeekendsIntervalMean$interval, stepsWeekendsIntervalMean$intervalMean, 
     type = "l", xlab = "5-minutes interval", ylab = "Mean number of steps in the    interval
     across all weekend", main = "Weekend")
```